---
output: github_document
bibliography: inst/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tempo

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/tempo)](https://CRAN.R-project.org/package=tempo)
[![R-CMD-check](https://github.com/joeroe/tempo/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/joeroe/tempo/actions/workflows/R-CMD-check.yaml)
[![Test coverage](https://codecov.io/gh/joeroe/tempo/graph/badge.svg)](https://app.codecov.io/gh/joeroe/tempo)
<!-- badges: end -->

**tempo** provides a formal representation of intervals between two points in time (periods) and the potential logical relations between them in R.

## Installation

You can install the development version of tempo from GitHub with the [pak](https://pak.r-lib.org/) package:

``` r
pak::pak("joeroe/tempo")
```

## Usage

Create intervals from numeric start and end dates:

```{r eg-interval}
library("tempo")

intv1 <- interval(1500, 1900)
intv2 <- interval(1800, 1950)
```

Or from numeric dates with an era label:

```{r eg-interval-era}
interval(1200, 800, "BCE")
```

Test whether two intervals are contemporary with each other:

```{r eg-contemporary_with}
contemporary_with(intv1, intv2)
```

tempo supports fifteen types of temporal relations following Levy's typology [@LevyEtAl2021; @Levy2025].
See `?relations` for a list.

By default, comparison of start and end points is inclusive (i.e. using `>=` and `<=`).
`strict = TRUE` enables the 'strict' variants of each relation (i.e. using `>` and `<`):

```{r eg-contemporary_with-strict}
intv3 <- interval(1900, 1950)

contemporary_with(intv1, intv3)

contemporary_with(intv1, intv3, strict = TRUE)
```

## References
