---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tempo

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/tempo)](https://CRAN.R-project.org/package=tempo)
[![R-CMD-check](https://github.com/joeroe/tempo/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/joeroe/tempo/actions/workflows/R-CMD-check.yaml)
[![Test coverage](https://codecov.io/gh/joeroe/tempo/graph/badge.svg)](https://app.codecov.io/gh/joeroe/tempo)
<!-- badges: end -->

***tempo*** is an R package that provides a formal representation of intervals between two points in time (periods) and the potential logical relations between them.

## Installation

You can install the development version of tempo from GitHub with the [remotes](https://remotes.r-lib.org/) package:

``` r
remotes::install_github("joeroe/tempo")
```

## Example

Test whether two periods, specified as numeric vectors of start and end dates, are contemporary with each other:

```{r}
library("tempo")

period1 <- c(1500, 1900)
period2 <- c(1800, 1950)

contemporary_with(period1, period2)
```

tempo supports fifteen types of temporal relations following Levy's typology (Levy et al. 2021 <https://doi.org/10.1016/j.jas.2020.105225>; Levy et al. in press).
See `?relations` for a list.

By default, comparison of start and end points is inclusive (i.e. using `>=` and `<=`).
`strict = TRUE` enables the 'strict' variants of each relation (i.e. using `>` and `<`):

```{r}
period3 <- c(1900, 1950)

contemporary_with(period1, period3)

contemporary_with(period1, period3, strict = TRUE)
```

