---
title: "Temporal intervals and relations"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Temporal intervals and relations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**tempo** provides a formal representation of intervals between two points in time (periods) and the logical relations between them.

Intervals are a foundational concept in chronological modelling across archaeology and other fields. 
Although R has several ways to represent spans of time (e.g. `Date` and `POSIXct`), these are based on the Gregorian calendar and are unsuited to non-Gregorian or deep-time applications. 
tempo instead represents intervals as vectors of start and end points on arbitrary time scales.
It is based on [vctrs](https://vctrs.r-lib.org), so the resulting S3 class is stable in data frames and tibbles, prints in a readable format, and behaves predictably in tidyverse workflows.
Optionally, the calendar era of year-based time scales can be explicitly specified via the [*era* package](https://era.joeroe.io), providing calendar-aware chronological operations.

Logical relations between temporal intervals were first studied by @Allen1983, with later archaeological adaptations by @Holst2001, @Holst2004, and the CIDOC-CRM standard [@CIDOCCRM2014]. 
They include relations like "x is before y", "x meets y", or "x overlaps y".
The package implements the typology developed by @LevyEtAl2021 and @Levy2025, which is a superset of these previous typologies.

This vignette introduces the main features of the package: constructing and inspecting temporal intervals, performing set operations on them, and testing the logical relations between them.

```{r setup}
library("tempo")
```

## Temporal intervals

The `interval()` function creates vectors of temporal intervals from numeric start (earliest) and end (latest) points:

```{r eg-interval-numeric}
interval(c(10, 20, 30), c(20, 30, 40))
```

To specify the calendar eras, pass an era label via the `era` argument:

```{r eg-interval-era}
interval(1200, 800, "BCE")
```

Or use `era::yr()` vectors directly:

```{r eg-interval-yr}
interval(era::yr(c(100, 200), "BP"), era::yr(c(50, 100), "BP"))
```

See the [era package vignette](https://era.joeroe.io/articles/era.html) for details on working with calendar eras.

Making the era explicitly is especially useful for backwards counted like BC(E) or Before Present, allowing for chronologically-aware arithmetic.
For example, `intv_duration()` takes into account the counting direction when calculating the length of each interval:

```{r eg-intv-duration}
x <- interval(1200, 800, "BCE")
intv_duration(x)
```

### Set operations

Two or more intervals can be combined using set operations.

`intv_union()` returns the bounding interval across all inputs:

```{r eg-intv-union}
a <- interval(10, 30)
b <- interval(20, 40)
intv_union(a, b)
```

`intv_intersection()` returns the overlapping region:

```{r eg-intv-intersection}
intv_intersection(a, b)
```

## Temporal relations

A temporal relation is a mathematical object describing the relationship between two temporal intervals, defined as a formal function of the intervals' four endpoints (the start and end of each). They avoid the ambiguities of natural language descriptions — for example, statements about whether two phases were "contemporary" can be read in several different ways, whereas a formal relation has a single, precise meaning. In chronological modelling they serve as an exact vocabulary, enabling computational analysis and consistent comparison of chronological claims across studies. The remainder of this section introduces the typology of such relations implemented in tempo @LevyEtAl2021; @Levy2025.

All relation functions share the signature `fn(x, y, strict = FALSE)` and accept `interval` objects, two-element numeric vectors, or lists of two-element numeric vectors.

By default, comparisons are **inclusive** (using `>=` and `<=`). Setting `strict = TRUE` uses **strict** comparisons (using `>` and `<`), which affects relations that involve equality of endpoints. For example, two intervals sharing an endpoint are `contemporary_with()` each other by default, but not under strict comparison:

```{r eg-relations}
intv1 <- interval(1500, 1900)
intv2 <- interval(1800, 1950)
intv3 <- interval(1900, 1950)

contemporary_with(intv1, intv2)

# Inclusive (default): intervals sharing an endpoint are contemporary
contemporary_with(intv1, intv3)

# Strict: intervals must overlap in their interiors
contemporary_with(intv1, intv3, strict = TRUE)
```

The package provides 24 functions for testing temporal relations, listed in the table below in the order of Levy's typology.

| Type | Relation | tempo function | Definition |
|------|----------|----------------|------------|
| Start–end order | Starts before or at end of | `starts_before_end_of()` | beg(x) ≤ end(y) |
| Start–end order | Ends after or at start of | `ends_after_start_of()` | end(x) ≥ beg(y) |
| Start order | Starts before or at start of | `starts_before_start_of()` | beg(x) ≤ beg(y) |
| Start order | Starts after or at start of | `starts_after_start_of()` | beg(x) ≥ beg(y) |
| End order | Ends before or at end of | `ends_before_end_of()` | end(x) ≤ end(y) |
| End order | Ends after or at end of | `ends_after_end_of()` | end(x) ≥ end(y) |
| Disjunction | Ends before or at start of | `ends_before_start_of()` | end(x) ≤ beg(y) |
| Disjunction | Starts after or at end of | `starts_after_end_of()` | beg(x) ≥ end(y) |
| Sequence | Meets | `meets()` | end(x) = beg(y) |
| Sequence | Met by | `met_by()` | beg(x) = end(y) |
| Contemporaneity | Contemporary with | `contemporary_with()` | end(x) ≥ beg(y) AND beg(x) ≤ end(y) |
| Start inclusion | Starts during | `starts_during()` | beg(y) ≤ beg(x) ≤ end(y) |
| Start inclusion | Includes start of | `includes_start_of()` | beg(x) ≤ beg(y) ≤ end(x) |
| End inclusion | Ends during | `ends_during()` | beg(y) ≤ end(x) ≤ end(y) |
| End inclusion | Includes end of | `includes_end_of()` | beg(x) ≤ end(y) ≤ end(x) |
| Equal start | Starts with | `starts_with()` | beg(x) = beg(y) |
| Equal end | Ends with | `ends_with()` | end(x) = end(y) |
| Overlap | Overlaps before | `overlaps_before()` | beg(x) ≤ beg(y) ≤ end(x) ≤ end(y) |
| Overlap | Overlaps after | `overlaps_after()` | beg(y) ≤ beg(x) ≤ end(y) ≤ end(x) |
| Inclusion | Includes | `includes()` | beg(x) ≤ beg(y) AND end(x) ≥ end(y) |
| Inclusion | Included in | `included_in()` | beg(x) ≥ beg(y) AND end(x) ≤ end(y) |
| Beginning | Begins | `begins()` | beg(x) = beg(y) AND end(x) ≤ end(y) |
| Beginning | Begun by | `begun_by()` | beg(x) = beg(y) AND end(x) ≥ end(y) |
| Ending | Ends | `ends()` | end(x) = end(y) AND beg(x) ≥ beg(y) |
| Ending | Ended by | `ended_by()` | end(x) = end(y) AND beg(x) ≤ beg(y) |
| Equality | Equals | `equal_to()` | beg(x) = beg(y) AND end(x) = end(y) |

Adapted from [-@Levy2025, Table 5]. The definitions above are for the inclusive variants; set `strict = TRUE` for the exclusive variants.

## References
